---
title: "Health and Economic Impact of Weather Events in the US"
author: "Vadim Galenchik"
date: "Tuesday, August 18, 2015"
output: html_document
---

##Synonpsis

In this article, the impact of different weather events on public health and economy is  analyzed. This investigation is based on the storm database collected from the U.S. National Oceanic and Atmospheric Administration’s (NOAA) from 1950 - 2011. The estimates of fatalities, injuries, property and crop damage are used to figure out what types of severe weather events are most harmful to the population health and the economy. This investigation shows that excessive heat and tornado are most harmful for population health, while flood, drought, and hurricane/typhoon have the greatest economic consequences.

##Data Processing

##Basic settings

Load libraries and get the R version info.

```{r}
library(ggplot2)
library(plyr)
require(gridExtra)
options(scipen = 999)  # Turn off scientific notations for numbers

R.version
```


Download archive and read .csv file 
```{r}
zipFile <- "data.csv.bz2"

if (!zipFile %in% dir()) 
{
    download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = zipFile)
}

if (!"stormData" %in% ls()) 
{
    stormData <- read.csv(bzfile(zipFile))
}
```

Get the total number of records and variables

```{r}
dim(stormData)
```

Since event type format is not specified it is necessary to do some preprocessing to merge the same event records. For example, there are events with types "small hail", "Small Hail" and "SMALL HAIL". It is obviosly that these are the same type of event. In some cases there are punctuation characters in the event names. So to remove this duplications it is necessary to convert the case and substitute punctuation character with spaces. Of cause this simple preprocessing does not guarantied that all duplications are merged but results get better.

```{r}
# translate all letters to lowercase
stormData$EVTYPE <- tolower(stormData$EVTYPE)
# replace all punct. characters with a space
stormData$EVTYPE <- gsub("[[:blank:][:punct:]+]", " ", stormData$EVTYPE)
length(unique(stormData$EVTYPE))
```

## Dangerous Events with respect to Population Health and Economic Effects

To find the event types that are most harmful to population health, the number of casualties are aggregated by the event type.

To analyze the impact of weather events on the economy, available property  and crop damage reportings/estimates are used.

In the raw data, the  damage estimates are represented with two fields, a  mantissa (PROPDMG for property damage and CROPDMG for crop damage) and the exponent (PROPDMGEXP for property damage and CROPDMGEXP for crop damage).  It should be noted that exponent may be presented as a number or as letter  mnemonic (h - hundred, k - thousand, m - million, b - billion). The first step in the economic effects analysis is to calculate the property and crop damage for each event.

```{r}
getExp <- function(e) 
{
    # h -> 10^2, k -> 10^3, m -> 10^6, b -> 10^9
    
    m <- 0
    if (e %in% c('h', 'H'))
        m <- (2)
    else if (e %in% c('k', 'K'))
        m <- (3)
    else if (e %in% c('m', 'M'))
        m <- (6)
    else if (e %in% c('b', 'B'))
        m <- (9)
    else if (!is.na(as.numeric(e)))
        m <- (as.numeric(e))
    else if (e %in% c('', '-', '?', '+'))
        m <- (0)
    else 
    {
        stop("Invalid exponent value.")
    }
    return (10**m)
}

stormData$PROPDMG_VAL <- stormData$PROPDMG * sapply(stormData$PROPDMGEXP, getExp)
stormData$CROPDMG_VAL <- stormData$CROPDMG * sapply(stormData$CROPDMGEXP, getExp)
```
Now the data may be aggregated by the event type.

```{r}
harm <- ddply(stormData, .(EVTYPE), summarize,
                    killed = sum(FATALITIES),
                    injured = sum(INJURIES),
                    property = sum(PROPDMG_VAL),
                    crop = sum(CROPDMG_VAL))

```

Find events that caused most death and injury.
```{r}
fatalEvents <- head(data.frame(type=harm$EVTYPE, killed = harm$killed)
                     [order(harm$killed, decreasing = T), ], 15)
fatalEvents

injuryEvents <- head(data.frame(type=harm$EVTYPE, injured = harm$injured)
                      [order(harm$injured, decreasing = T), ], 15)
injuryEvents
```
The following plot shows top dangerous weather event types.
```{r}
fatalitiesPlot <- qplot(type, data = fatalEvents, weight = killed, geom = "bar") + 
    aes(x=reorder(type, killed)) +
    scale_y_continuous("Number of Fatalities") +  
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    xlab("Severe Weather Type") +
    ggtitle("Total Fatalities by Severe Weather\n Events in the U.S.\n from 1950 - 2011")

injuriesPlot <- qplot(type, data = injuryEvents, weight = injured, geom = "bar") +
    aes(x=reorder(type, injured))+
    scale_y_continuous("Number of Injuries") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    xlab("Severe Weather Type") +
    ggtitle("Total Injuries by Severe Weather\n Events in the U.S.\n from 1950 - 2011")

grid.arrange(fatalitiesPlot, injuriesPlot, ncol = 2)
```

This figure shows that tornadoes cause most number of deaths and injuries among all event types. There are more than 5,000 deaths and more than 10,000 injuries in the last 60 years in the US, caused tornadoes. The other event types that are most dangerous with respect to population health are excessive heat and flash floods.

Find events that caused most economical impact.
```{r}
propertyDamage <- head(data.frame(type=harm$EVTYPE, property = harm$property)
                       [order(harm$property, decreasing = T), ], 10)
format(propertyDamage, big.mark=",")

cropDamage <- head(data.frame(type=harm$EVTYPE, crop = harm$crop)
                   [order(harm$crop, decreasing = T), ], 10)
format(cropDamage, big.mark=",")
```

The following plot shows top dangerous for economy weather event types.

```{r}
propertyPlot <- qplot(type, data = propertyDamage, weight = property/1e12, geom = "bar") + 
    aes(x=reorder(type, property)) +
    scale_y_continuous("Property Damage, trillon USD") +  
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    xlab("Severe Weather Type") +
    ggtitle("Total Property Damage by \n Severe Weather Events \n in the U.S. from 1950 - 2011")

cropPlot <- qplot(type, data = cropDamage, weight = crop/1e9, geom = "bar") +
    aes(x=reorder(type, crop))+
    scale_y_continuous("Crop Damage, billion USD") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    xlab("Severe Weather Type") +
    ggtitle("Total Crop Damage by \n Severe Weather Events \n in the U.S. from 1950 - 2011")

grid.arrange(propertyPlot, cropPlot, ncol = 2)
```

This figure shows that flash floods and thunderstorm winds cause the largest property damages. 

The most severe weather event in terms of crop damage is the drought. In the last half century, it has caused about 14 billion dollars damage. Other severe crop-damage-causing events are floods and hails.

It should be noted that since the available data are not tidy, types "flood", "river flood", "flash flood", etc are separate observation. They should be merged for more accurate data-driven conclusions.
